---
title: "STAT 508 Data Analysis Assignment 7"
author: "Sebastian Bautista"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment='>')

SR = read.table("ABBREV.txt", header=F, row.names=1, sep="^", quote="~")
SR = na.omit(SR) # remove rows with missing values
SR = SR[row.names(SR) != "13352",] # remove "duplicate" entry
row.names(SR) = SR[,1] # set more meaningful row names
SR = SR[,-1]
names(SR) = c("Water_(g)", "Energ_Kcal", "Protein_(g)", "Lipid_Tot_(g)", "Ash_(g)", "Carbohydrt_(g)", "Fiber_TD_(g)", "Sugar_Tot_(g)", "Calcium_(mg)", "Iron_(mg)", "Magnesium_(mg)", "Phosphorus_(mg)", "Potassium_(mg)", "Sodium_(mg)", "Zinc_(mg)", "Copper_(mg)", "Manganese_(mg)", "Selenium_(µg)", "Vit_C_(mg)", "Thiamin_(mg)", "Riboflavin_(mg)", "Niacin_(mg)", "Panto_Acid_(mg)", "Vit_B6_(mg)", "Folate_Tot_(µg)", "Folic_Acid_(µg)", "Food_Folate_(µg)", "Folate_DFE_(µg)", "Choline_Tot_(mg)", "Vit_B12_(µg)", "Vit_A_IU", "Vit_A_RAE", "Retinol_(µg)", "Alpha_Carot_(µg)", "Beta_Carot_(µg)", "Beta_Crypt_(µg)", "Lycopene_(µg)", "Lut+Zea_(µg)", "Vit_E_(mg)", "Vit_D_µg", "Vit_D_IU", "Vit_K_(µg)", "FA_Sat_(g)", "FA_Mono_(g)", "FA_Poly_(g)", "Cholestrl_(mg)", "GmWt_1", "GmWt_Desc1", "GmWt_2", "GmWt_Desc2", "Refuse_Pct")
df = SR[,c(1:46)] # restrict to just the nutrient variables
```

# 0. Introduction

This week, we're applying PCA to USDA food composition data. After performing PCA I plot the proportion of variance explained in order to decide on an optimal number of components. I then examine the first three components in detail and attempt to make sense of the PC loadings by coming up with real life examples of foods that fit the descriptions.

# 1. Data

We're looking at a food composition data set with 2223 observations and 46 variables, where each observation is a specific food item and each variable is the mean nutrient value per 100g of the edible portion of the food, for example protein content in grams or vitamin C content in milligrams. All variables are either `num` or `int` (no factors or strings). The structure of the data set can be found in **figure 1**. 

# 2. Analyses

**Figure 2** contains a summary of performing PCA on the data. The summary reports the standard deviation, proportion of variance explained (PVE), and cumulative PVE. The first component explains 15.9% of the variance in the data, the second component 9.4%, the third component 9.1%, and so on.

PVE and cumulative PVE for each principal component are easier to show through plots. A combined scree plot and cumulative PVE plot can be found in **figure 3.1**, with PVE by component represented in blue and cumulative PVE in orange. If our goal is to explain a certain percentage of the variance, we can eyeball the cumulative PVE, but in this exercise I base my criteria on the scree plot slope leveling off. Since it's hard to eyeball an elbow in this figure, a zoomed-in scree plot is shown in **figure 3.2**.

Judging from the zoomed-in scree plot, it looks like the slope levels off after 7 principal components. At PC7, the cumulative PVE is 56.47%, yielding an average of about 8% per PC, which seems like a reasonable stopping point. Put another way, we only need around 15% of the components to explain over 56% of the variance in the data. 

*Are the first few principal components capturing any interpretable combinations of the nutrients?*

I report the first three principal components' loadings in **figures 4.1 through 4.3**. The first principal component is interesting - only water, alpha carotene, and lycopene are positive (with the latter two very close to zero), and the rest of the variables are negative. This could simply mean that water displaces nutrients in food, or equivalently, the more water in a food item, the less space there is for things like sugars, protein, and vitamins. Some foods that have high water content and relatively low nutrition content may be fruits and vegetables like watermelons and tomatoes. Conversely, dried versions of foods (raisins, prunes, etc.) are denser than their fresh variants and should therefore have more nutrients per 100g.

The loadings for the second PC are in **figure 4.2**. In this case, protein, fats, cholesterol, calories, and some water soluble vitamins (B12, B6, Niacin) are positive, while beta carotene, fiber, carbs, and some fat soluble vitamins (A, K) are negative. Generally speaking, most meats fit into this description. An example of something on the opposite side of the spectrum might be carrots, parsnips, or other root vegetables with natural sugars - vitamin A, fiber, carbs, and beta carotene fit the bill.

Finally, the loadings for the third PC are in **figure 4.3**. Here, water, protein, and vitamins are positive, while calories, fats, and carbs are negative. This PC seems to broadly represent "healthy" and "unhealthy" foods - leafy greens and lean fish and meats on one hand, and potato chips and other fried foods on the other. 

# 3. Plots and Tables

```{r}
str(df)
```
\begin{center}\textbf{Figure 1 - Structure of the SRp data set}\end{center}
\hrulefill

```{r}
pr.out = prcomp(df, scale=T)

summary(pr.out)
```
\begin{center}\textbf{Figure 2 - PCA summary}\end{center}
\hrulefill

```{r}
pr.var = pr.out$sdev^2
pve = pr.var/sum(pr.var)

plot(pve, xlab='Principal Component', ylab='Proportion of Variance Explained', 
     ylim=c(0,1), type='b', col='blue', main='Scree and cumulative PVE plot')
points(cumsum(pve), col='darkorange')

text(40, 0.1, label='PVE by component', col='blue')
text(40, 0.9, label='Cumulative PVE', col='darkorange')
```
\begin{center}\textbf{Figure 3.1 - PVE plots}\end{center}
\hrulefill

```{r}
# hard to see the elbow in that plot. scale to max(pve)
plot(pve, xlab='Principal Component', ylab='Proportion of Variance Explained', 
     ylim=c(0,max(pve)), type='b', col='blue', main='Scree plot, zoomed')
# slope levels off after 7, so elbow at 7
# cumulative PVE is 56.47%
```
\begin{center}\textbf{Figure 3.2 - Scree plot, zoomed}\end{center}
\hrulefill

```{r}
# principal component loading vectors (weighting of variables), len=p=46
pr.out.rotation = round(pr.out$rotation, 3)

sort(pr.out.rotation[,1])
```
\begin{center}\textbf{Figure 4.1 - Loadings for PC1}\end{center}
\hrulefill

```{r}
# second principal component and its loadings
sort(pr.out.rotation[,2])
```
\begin{center}\textbf{Figure 4.2 - Loadings for PC2}\end{center}
\hrulefill

```{r}
# third principal component and its loadings
sort(pr.out.rotation[,3])
```
\begin{center}\textbf{Figure 4.3 - Loadings for PC3}\end{center}
\hrulefill

# 4. Conclusions

In this exercise, I performed PCA on food composition data, chose an optimal number of principal components, and examined the loadings for the first few PCs in order to make sense of relationships between variables. I came up with examples of foods whose differences can be explained in terms of the loadings: the first PC looked like a wet-dry axis, the second seemed to be meats-vegetables, and the third could be described as healthy-unhealthy. 

# 5. Appendix

```{r ref.label=knitr::all_labels(), echo=T, eval=F}
```