---
title: "STAT 508 Data Analysis Assignment 10"
author: "Sebastian Bautista"
output:
  pdf_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(comment='>', echo=F)
options(digits=2)

library(ISLR)
library(MASS)
df = Weekly
names(df) = tolower(names(df))

train = (df$year < 2009)
X = scale(df[,-9]) # SCALING DF

Xtest = df[!train,]
Xtrain = df[train,]
ytest = df$direction[!train]
ytrain = df$direction[train]
```

# 0. Introduction

*Repeat step 4 from last week's prompts using kNN with k = 1.*

*How does kNN compare with logistic regression, LDA, and QDA results you obtained last week? (You can reference last week's report directly. No need to re-run the analyses, unless you want to improve on them.)*

*Experiment with different combinations of predictors, including possible transformations and interactions. Experiment with values for k in the kNN classifier. Report the variables, classifier, and associated confusion matrix that appears to provide the best results on the held out data.*

# 1. Data

# 2. Analyses

# 3. Plots and Tables

# 4. Conclusions

# 5. Appendix

```{r ref.label=knitr::all_labels(), echo=T, eval=F}
```
