---
title: "STAT 508 Data Analysis Assignment 4"
author: "Sebastian Bautista"
output:
  pdf_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(comment='>')
```

# 0. Introduction

This week, we're returning to the Boston data set from the `MASS` library. In this analysis, our response variable is `crim`, the per capita crime rate by town.

# 1. Data

# 2. Analyses

## 2.1 

*Q: Conduct an exploratory data analysis to identify any data peculiarities. Any possible colinearities? Any transformations needed (e.g., logarithm transformation)?*

## 2.2

*Q: Perform best subset selection, the lasso, and ridge regression. Present and discuss results for the approaches that you consider.*

## 2.3

*Q: Propose a model (or set of models) that seem to perform well on this dataset, and justify your proposal. Make sure that you are evaluating model performance using validation set error and/or cross-validation, as shown in ISLR Sections 6.5 and 6.6.*

## 2.4

*Q: Does your chosen model involve all of the features in the data set? Why or why not?*

# 3. Plots and Tables

```{r}
library(MASS)
df = Boston
```

```{r}
# Scratch work, ideas, etc.

## CHECK MISSINGS?
sapply(df, function(x) sum(is.na(x)))

## TRANSFORM DEP VAR?
hist(df$crim) # positive outliers. skew
df$crim.log = log(df$crim) # new variable. log of dep var
hist(df$crim.log) # looks better than before
```



# 4. Conclusions

# 5. Appendix

```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```