---
title: "STAT 508 Data Analysis Assignment 8"
author: "Sebastian Bautista"
output:
  pdf_document: default
---

```{r setup, include=FALSE, cache=T}
knitr::opts_chunk$set(comment='>')

library(readxl)
eretail = read_excel("Online Retail.xlsx")
dim(eretail)
names(eretail)

eretail = eretail[eretail$Country != "Unspecified",] # remove 'unspecified' country
eretail = eretail[eretail$Quantity > 0,]             # remove returns/cancellations

IDtab = table(eretail$Country, eretail$CustomerID)   # crosstab country by customer ID
IDtab = apply(IDtab >0, 2, sum)                      # is any customer ID duplicated across countries?
duplicateIDs = names(IDtab[IDtab > 1])               # duplicate IDs to clean up
eretail = eretail[!is.element(eretail$CustomerID, duplicateIDs),]
rm(IDtab)

eretail$InvoiceMth = substr(eretail$InvoiceDate, 1, 7)         # extract month of invoice
eretail = eretail[as.vector(eretail$InvoiceMth) != "2011-12",] # remove December 2011 as it only covers first week

eretail$Amount = eretail$Quantity * eretail$UnitPrice           # compute amount per invoice item

eaggr = aggregate(Amount~Country+CustomerID, data=eretail, sum) # compute aggregate amount spent per customer
row.names(eaggr) = eaggr$CustomerID
eaggr = eaggr[,-2]
eaggr = cbind(eaggr, aggregate(InvoiceMth~CustomerID, data=eretail, min)[,-1]) # 1st month of customer interaction
names(eaggr)[3] = "FirstMth"
eaggr = cbind(eaggr, aggregate(InvoiceMth~CustomerID, data=eretail, max)[,-1]) # last month of cust. interaction
names(eaggr)[4] = "LastMth"

# relabel months and compute duration of customer interaction
levels(eaggr$FirstMth) = 1:12
levels(eaggr$LastMth) = 1:12
eaggr$FirstMth = as.numeric(as.vector(eaggr$FirstMth))
eaggr$LastMth = as.numeric(as.vector(eaggr$LastMth))
eaggr$Months = eaggr$LastMth - eaggr$FirstMth + 1

eaggr = cbind(eaggr, apply( table(eretail$CustomerID, eretail$InvoiceMth) , 1, sum ) )
names(eaggr)[6] = "Purchases"

# Some useful statistics (which you may or may not decide to use)
eaggr$Amount.per.Purchase = eaggr$Amount / eaggr$Purchases
eaggr$Purchases.per.Month = eaggr$Purchases / eaggr$Months
eaggr$Amount.per.Month = eaggr$Amount / eaggr$Months

eaggr[1:30,]

df = eaggr
colnames(df) = tolower(colnames(df))
```

# 0. Introduction

This week, we're using clustering techniques on sales data from a UK company. The original data is at the transaction level, but we aggregate to the customer level and create a few more variables in order to do our analysis. 

# 1. Data

# 2. Analyses

*Using cluster analysis, your aim is to identify whether this company's consumers can be segmented meaningfully by the recency and duration of their interaction with the company, the frequency of their purchases/orders, and the amount they spent (amounts are in Sterling pounds). Further segmentation by country would also be very helpful for marketing purposes.*

Recency - `lastmth`
Duration - `months`
Frequency - `purchases`, `purchases.per.month`
Amount - `amount`, `amount.per.purchase`, `amount.per.month`

*For the data source, documentation, and an interesting study using a variant of this data, please see https://archive.ics.uci.edu/ml/datasets/Online%20Retail*

# 3. Plots and Tables

```{r}
cat('The data contains', length(unique(df$country)), 'unique countries\n')
table(df$country)

# create `intl` dummy variable because of predominance of UK customers
# and to coerce `country` into something numeric and sensible
df$intl = 0
mask = which(df$country != 'United Kingdom')
df$intl[mask] = 1
df$country = NULL

# scale and print summary
df = scale(df)
row.names(df) = NULL
summary(df)
```

\begin{center}\textbf{Figure X}\end{center}

\hrulefill

```{r}
# K-means clustering, k=2
set.seed(7)
km.2 = kmeans(df, 2, nstart=50)

cat('K-means clustering with', length(km.2$size), 'clusters of size', km.2$size[1], ',', km.2$size[2])

# cluster means
cat('\n\nCluster means:\n')
round(km.2$centers, 4) 
# within cluster sum of squares - variance in data set explained by clustering
# k-means minimizes within group dispersion and maximizes between group dispersion
# by assigning the obs to these clusters rather than n clusters,
# we achieve a reduction in sum of squares of 17.3%
cat('\nBetween ss / Total ss:\n')
round(km.2$betweenss/km.2$totss, 4)
```

\begin{center}\textbf{Figure X}\end{center}

\hrulefill

```{r}
# K-means clustering, k=3
km.3 = kmeans(df, 3, nstart=50)

cat('K-means clustering with', length(km.3$size), 'clusters of size', 
    km.3$size[1], ',', km.3$size[2], ',', km.3$size[3])

# cluster means
cat('\n\nCluster means:\n')
round(km.3$centers, 4) 

# within cluster sum of squares - variance in data set explained by clustering
cat('\nBetween ss / Total ss:\n')
round(km.3$betweenss/km.3$totss, 4)
```

\begin{center}\textbf{Figure X}\end{center}

\hrulefill

```{r}
# K-means clustering, k=4
km.4 = kmeans(df, 4, nstart=50)

cat('K-means clustering with', length(km.4$size), 'clusters of size', 
    km.4$size[1], ',', km.4$size[2], ',', km.4$size[3], ',', km.4$size[4])

# cluster means
cat('\n\nCluster means:\n')
round(km.4$centers, 4) 

# within cluster sum of squares - variance in data set explained by clustering
cat('\nBetween ss / Total ss:\n')
round(km.4$betweenss/km.4$totss, 4)
```

\begin{center}\textbf{Figure X}\end{center}

\hrulefill

```{r}
# K-means clustering, k=5
# might stop here because we get a singleton observation
km.5 = kmeans(df, 5, nstart=50)

cat('K-means clustering with', length(km.5$size), 'clusters of size', 
    km.5$size[1], ',', km.5$size[2], ',', km.5$size[3], ',', km.5$size[4], ',', km.5$size[5])

# cluster means
cat('\n\nCluster means:\n')
round(km.5$centers, 4) 

# within cluster sum of squares - variance in data set explained by clustering
cat('\nBetween ss / Total ss:\n')
round(km.5$betweenss/km.5$totss, 4)
```

\begin{center}\textbf{Figure X}\end{center}

\hrulefill

```{r}
# K-means clustering, k=6
# too much? might delete and move on to hierarchical
km.6 = kmeans(df, 6, nstart=50)

cat('K-means clustering with', length(km.6$size), 'clusters of size', 
    km.6$size[1], ',', km.6$size[2], ',', km.6$size[3], ',', km.6$size[4], ',', km.6$size[5],
    ',', km.6$size[6])

# cluster means
cat('\n\nCluster means:\n')
round(km.6$centers, 4) 

# within cluster sum of squares - variance in data set explained by clustering
cat('\nBetween ss / Total ss:\n')
round(km.6$betweenss/km.6$totss, 4)
```

\begin{center}\textbf{Figure X}\end{center}

\hrulefill

```{r}
# hierarchical clustering
# complete and average look ok but single looks terrible
# will show it here, but compare k-means to complete and average later
df.dist = dist(df)

hcc = hclust(df.dist)
hca = hclust(df.dist, method='average')
hcs = hclust(df.dist, method='single')
hcce = hclust(df.dist, method='centroid')

plot(hcc, main="Complete Linkage", xlab="", ylab="", sub="")
plot(hca, main="Average Linkage", xlab="", ylab="", sub="")
plot(hcs, main="Single Linkage", xlab="", ylab="", sub="")
plot(hcce, main="Centroid Linkage", xlab="", ylab="", sub="")
```

\begin{center}\textbf{Figure X}\end{center}

\hrulefill

```{r}
hcc.clusters.10 = cutree(hcc, 10)
table(hcc.clusters.10)

hcc.clusters.9 = cutree(hcc, 9)
table(hcc.clusters.9)

hcc.clusters.8 = cutree(hcc, 8)
table(hcc.clusters.8)

hcc.clusters.7 = cutree(hcc, 7)
table(hcc.clusters.7)

hcc.clusters.6 = cutree(hcc, 6)
table(hcc.clusters.6)

hcc.clusters.5 = cutree(hcc, 5)
table(hcc.clusters.5)

hcc.clusters.4 = cutree(hcc, 4)
table(hcc.clusters.4)

hcc.clusters.3 = cutree(hcc, 3)
table(hcc.clusters.3)

# looks pretty weird. probably not the way to go
# anything between k=3 and k=10 leads to one huge cluster and lots of small ones
```

\begin{center}\textbf{Figure X}\end{center}

\hrulefill

```{r}
hca.clusters.10 = cutree(hca, 10)
table(hca.clusters.10)

hca.clusters.9 = cutree(hca, 9)
table(hca.clusters.9)

hca.clusters.8 = cutree(hca, 8)
table(hca.clusters.8)

hca.clusters.7 = cutree(hca, 7)
table(hca.clusters.7)

hca.clusters.6 = cutree(hca, 6)
table(hca.clusters.6)

hca.clusters.5 = cutree(hca, 5)
table(hca.clusters.5)

hca.clusters.4 = cutree(hca, 4)
table(hca.clusters.4)

hca.clusters.3 = cutree(hca, 3)
table(hca.clusters.3)
# same as above
```

\begin{center}\textbf{Figure X}\end{center}

\hrulefill

```{r}
hcs.clusters.10 = cutree(hcs, 10)
table(hcs.clusters.10)

hcs.clusters.9 = cutree(hcs, 9)
table(hcs.clusters.9)

hcs.clusters.8 = cutree(hcs, 8)
table(hcs.clusters.8)

hcs.clusters.7 = cutree(hcs, 7)
table(hcs.clusters.7)

hcs.clusters.6 = cutree(hcs, 6)
table(hcs.clusters.6)

hcs.clusters.5 = cutree(hcs, 5)
table(hcs.clusters.5)

hcs.clusters.4 = cutree(hcs, 4)
table(hcs.clusters.4)

hcs.clusters.3 = cutree(hcs, 3)
table(hcs.clusters.3)
# again same as above and even worse, with tons of singletons
```

\begin{center}\textbf{Figure X}\end{center}

\hrulefill

```{r}
hcce.clusters.10 = cutree(hcce, 10)
table(hcce.clusters.10)

hcce.clusters.9 = cutree(hcce, 9)
table(hcce.clusters.9)

hcce.clusters.8 = cutree(hcce, 8)
table(hcce.clusters.8)

hcce.clusters.7 = cutree(hcce, 7)
table(hcce.clusters.7)

hcce.clusters.6 = cutree(hcce, 6)
table(hcce.clusters.6)

hcce.clusters.5 = cutree(hcce, 5)
table(hcce.clusters.5)

hcce.clusters.4 = cutree(hcce, 4)
table(hcce.clusters.4)

hcce.clusters.3 = cutree(hcce, 3)
table(hcce.clusters.3)
# no improvement. go with kmeans
```

\begin{center}\textbf{Figure X}\end{center}

\hrulefill

```{r}
# example of using hclust to visualize k=5 clusters for `hcc`
plot(hcc, labels=F, main="Complete Linkage", xlab="", ylab="", sub="")
rect.hclust(hcc, k=5)
```

\begin{center}\textbf{Figure X}\end{center}

\hrulefill

# 4. Conclusions

# 5. Appendix

```{r ref.label=knitr::all_labels(), echo=T, eval=F}
```