---
title: "EDA"
author: "Sebastian"
date: "April 21, 2019"
output: html_document
---

```{r setup, include=FALSE}
df = read.csv('../data/OnlineNewsPopularity.csv')
```

```{r}
str(df)
```

Structure of the data, almost completely numeric. `shares` is the target variable, and all but `url` and `timedelta` are predictive of the target. `timedelta` is probably not that useful (days between the article publication and the dataset acquisition (non-predictive)) and `url` is probably useful just for pulling out the article name if we need it.

```{r}
sapply(df, function(x) sum(is.na(x)))
```

Column-wise missings. Nothing is missing.

```{r}
summary(df)
```

Summary of the data. Most likely not using all of these.
Which ones should we ignore? 

1. `timedelta` because it's non-predictive
2. Need to figure out the variables beginning with `kw` and what they mean
3. `LDA` variables probably beyond the scope of this course
4. `global_subjectivity` - opinions, allegations, desires, beliefs; 0:1
5. `global_sentiment_polarity` positive-negative polarity



```{r}
num = df[, sapply(df, is.numeric)]
dfcorr = cor(num)
sort(round(dfcorr, 4)[,'shares'])
# print rounded correlation coefficients with our target variable
# max of 11%, min of -5.9%
```

Pairwise correlation coefficients with `shares`.

```{r}
hist(df$shares)
hist(log(df$shares))
# lots of skew in `shares`
# might want to consider a log transformation as a dep var
df$shares.log = log(df$shares)
```

There's some serious right skew in `shares` and the log transformation looks closer to normal. Should probably try both when predicting.

```{r}
num = df[, sapply(df, is.numeric)]
dfcorr = cor(num)
sort(round(dfcorr, 4)[,'shares.log'])
```

Pairwise correlation with `shares.log`. Range seems larger. More informative?

