---
title: "STAT 508 Project 1"
author: "Sebastian Bautista"
output:
  pdf_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(comment='>')
library(ISLR)
library(GGally)
library(car)
library(leaps)
library(glmnet)
library(pls)
```

# 0. Introduction

For this project, we're looking at the College data set, aiming to predict the number of applications received by a college. 

# 1. Data

`College` is a cross-sectional data set where each observation is a US college from the 1995 issue of US News and World Report. Our response variable is `apps` and there are 17 other variables in the data. 

# 2. Analyses

1. *Examine the data using exploratory data analysis tools. Are there any features that would appear to jeopardize linear regression modeling? If so, address these issues before going any further.*

2. *Split the data into a training and a test set.*

3. *Using best subset selection, fit a least-squares regression model on the training set, and report the test error obtained.*

4. *Fit a ridge regression model on the training set, with lambda chosen by cross-validation. Report the test error obtained.*

5. *Repeat #4 with a lasso model instead of ridge regression, and report the test error obtained as well as the number of nonzero coefficient estimates.*

6. *Fit a PCR model on the training set, with M chosen by cross-validation. Report the test error as well as the value of M selected.*

7. *Repeat #6 with a PLS model instead of PCR.*

*Comment on the results obtained. What is the prediction accuracy? How do the different methods agree/disagree? Summarize your findings in a report, following the usual guidelines. Please pay special attention to your presentation, writing, and cohesiveness of analysis.* 


# 3. Plots and Tables

```{r}
# exploratory data analysis
df0 = College
# changing variable names to lower case
names(df0) = tolower(names(df0))
str(df0)
# checking for missings
cat(sum(is.na(df0)), 'missing values')
```
\begin{center}\textbf{Figure X}\end{center}
\hrulefill

```{r}
# look at first few obs
head(df0)
```
\begin{center}\textbf{Figure X}\end{center}
\hrulefill

```{r}
# p.undergrad, f.undergrad, enroll, accept, apps are positive skewed (mean>med)
# our sample is 72.7% private schools and 27.3% public
summary(df0)
```
\begin{center}\textbf{Figure X}\end{center}
\hrulefill

```{r}
# FIRST SIMPLE LINEAR MODEL
reg0 = lm(apps ~ ., data=df0)
summary(reg0)
# heteroskedasticity and non-normality in residuals but the fit is good overall
# Rutgers at New Brunswick – positive residuals indicate that we underestimate `apps`. In other words, they receive more `apps` than they should be according to the data. 
plot(reg0, which=1)
cat('MSE:',mean(summary(reg0)$residuals^2))
```
\begin{center}\textbf{Figure X}\end{center}
\hrulefill

```{r}
numdf = model.matrix(~., data=df0)[,-1]
# positive outliers and skew for apps, accept, p.undergrad, books, and expend
boxplot(scale(numdf), las=2, main='College data set, scaled')
```
\begin{center}\textbf{Figure X}\end{center}
\hrulefill

```{r}
# a lot of these pairs make sense that they would be highly collinear
# probably leaning towards feature selection because of this
# apps, our response, is highly correlated with accept, enroll, and f.undergrad
# apps, accept, enroll, and f.undergrad
# also talk about how private schools differ from public. higher tuition, fewer students
ggcorr(numdf, label=T)
```
\begin{center}\textbf{Figure X}\end{center}
\hrulefill

```{r}
# enroll and f.undergrad show high signs of multicollinearity
# accept, top10perc, and top25perc are also somewhat high
sort(vif(reg0), decreasing=T)
```
\begin{center}\textbf{Figure X}\end{center}
\hrulefill

```{r}
# possible obs to drop
# again, Rutgers at New Brunswick is pointed out
outlierTest(reg0)
```
\begin{center}\textbf{Figure X}\end{center}
\hrulefill

```{r}
# SECOND SIMPLE LINEAR MODEL
# dropping enroll, f.undergrad, and Rutgers at New Brunswick
df1 = data.frame(df0)
df1$enroll = NULL
df1$f.undergrad = NULL
remove.rows = c('Rutgers at New Brunswick')
df1 = df1[!(row.names(df1) %in% remove.rows), ]
reg1 = lm(apps ~ ., data=df1)
summary(reg1)
# again, weird residuals, but not as bad as before
plot(reg1, which=1)
cat('MSE:',mean(summary(reg1)$residuals^2))
# MSE is lower at 939537
```

\begin{center}\textbf{Figure X}\end{center}
\hrulefill

```{r}
# THIRD SIMPLE LINEAR MODEL
# log transforming the response variable
reg2 = lm(log(apps) ~ . - apps, data=df1)
summary(reg2)
# fit looks kinda bad and adj R^2 drops
plot(reg2, which=1)
cat('MSE:',mean(summary(reg2)$residuals^2))
```

\begin{center}\textbf{Figure X}\end{center}
\hrulefill

```{r}
# FOURTH SIMPLE LINEAR MODEL
# log transforming skewed predictors
# positive outliers and skew for apps, accept, p.undergrad, books, and expend
reg3 = lm(log(apps) ~ private + log(accept) + top10perc + top25perc + log(p.undergrad)
          + outstate + room.board + log(books) + personal + phd + terminal
          + s.f.ratio + perc.alumni + log(expend) + grad.rate, data=df1)
summary(reg3)
# fit looks much better 
plot(reg3, which=1)
cat('MSE:',mean(summary(reg3)$residuals^2))
```

\begin{center}\textbf{Figure X}\end{center}
\hrulefill

```{r}
# creating final dataset for the rest of the analyses
df1$apps.log = log(df1$apps)
df1$accept.log = log(df1$accept)
df1$p.undergrad.log = log(df1$p.undergrad)
df1$books.log = log(df1$books)
df1$expend.log = log(df1$expend)
df1$apps = NULL
df1$accept = NULL
df1$p.undergrad = NULL
df1$books = NULL
df1$expend = NULL
```

\begin{center}\textbf{Figure X}\end{center}
\hrulefill

```{r}
# BEST SUBSET SELECTION

df = data.frame(df1)

# train-test split
set.seed(90210)
train = sample(1:nrow(df), nrow(df)/2)
test = (-train)
# number of predictors p
nvmax = length(names(df)) - 1
bss = regsubsets(apps.log ~ ., data=df[train,], nvmax=nvmax)
test.mat = model.matrix(apps.log ~ ., data=df[test,])
val.errors = rep(NA,nvmax)
for(i in 1:nvmax){
coefi = coef(bss, id=i)
pred = test.mat[,names(coefi)] %*% coefi
val.errors[i] = mean((df$apps.log[test] - pred)^2)
}
num.vars = which.min(val.errors)
coef(bss, num.vars)
paste("The winning model contains", num.vars, "variables")
paste("The test set MSE for the model chosen by best subset selection is", round(val.errors[num.vars], 4))
# still the best. BSS is generally better in high signal-to-noise ratio regimes, so we can conclude
# that a lot of these variables are provide useful information when it comes to predicting `apps.log`
```
\begin{center}\textbf{Figure X - Best subset selection}\end{center}
\hrulefill

```{r}
# RIDGE
X = model.matrix(apps.log ~ ., df)[,-1]
y = df$apps.log
Xtrain = X[train,]
Xtest = X[test,]
ytrain = y[train]
ytest = y[test]
lambda = 10^seq(10, -4, length=5000)
cv.ridge = cv.glmnet(Xtrain, ytrain, alpha=0, lambda=lambda)
ridge.lambda = cv.ridge$lambda.min
ridge = glmnet(Xtrain, ytrain, alpha=0, lambda=lambda)
ridge.yhat = predict(ridge, s=ridge.lambda, newx=Xtest)
ridge.test.mse = mean((ridge.yhat - ytest)^2)
ridge.full = glmnet(X, y, alpha=0)
# Refit on whole data set
print(predict(ridge.full, type='coefficients', s=ridge.lambda))
paste("The value of lambda that returns the smallest CV error for ridge is", round(ridge.lambda, 3))
paste("The test set MSE for ridge is", round(ridge.test.mse, 4))
```
\begin{center}\textbf{Figure X - Ridge}\end{center}
\hrulefill

```{r}
# LASSO
cv.lasso = cv.glmnet(Xtrain, ytrain, alpha=1, lambda=lambda)
lasso.lambda = cv.lasso$lambda.min
lasso = glmnet(Xtrain, ytrain, alpha=1, lambda=lambda)
lasso.yhat = predict(lasso, s=lasso.lambda, newx=Xtest)
lasso.test.mse = mean((lasso.yhat - ytest)^2)
lasso.full = glmnet(X, y, alpha=1)
# Refit on whole data set
print(predict(lasso.full, type='coefficients', s=lasso.lambda))
paste("The value of lambda that yields the smallest CV error for lasso is", round(lasso.lambda, 3))
paste("The test set MSE for lasso is", round(lasso.test.mse, 4))
```
\begin{center}\textbf{Figure X - Lasso}\end{center}
\hrulefill

```{r}
# PCR
cv.pcr = pcr(apps.log ~ ., data=df, subset=train, scale=T, validation='CV')
cverr = MSEP(cv.pcr)$val[1,,]
ncomp = (which.min(cverr) - 1)[[1]]
pcr.yhat = predict(cv.pcr, Xtest, ncomp=ncomp)
pcr.mse = mean((pcr.yhat - ytest)^2)
pcr.full = pcr(y~X, scale=T, ncomp=ncomp)
cat('Optimal number of principal components: ', ncomp)
cat('Test MSE for pcr: ', round(pcr.mse, 4))
# 16 principal components = OLS
```
\begin{center}\textbf{Figure X - Principal Components Regression}\end{center}
\hrulefill

```{r}
# PLSR
cv.pls = plsr(apps.log ~ ., data=df, subset=train, scale=T, validation='CV')
cverr = MSEP(cv.pls)$val[1,,]
ncomp = (which.min(cverr) - 1)[[1]]
pls.yhat = predict(cv.pls, Xtest, ncomp=ncomp)
pls.mse = mean((pls.yhat - ytest)^2)
pls.full = plsr(y~X, scale=T, ncomp=ncomp)
cat('Optimal number of components: ', ncomp)
cat('Test MSE for pls: ', round(pls.mse, 4))
```
\begin{center}\textbf{Figure X - Partial Least Squares Regression}\end{center}
\hrulefill

# 4. Conclusions

# 5. Appendix

```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```